Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-10-09 Writing/NatureWater/3rdRoundRevision/replies/reply_to_reviewers_R3.rmd

---
output:
  pdf_document: 
    keep_tex: true
  html_document: default
---

# Reply to Reviewer 1

I have read the second revision of this manuscript and the authors' responses to my last review. I think the authors have done an excellent job in responding to the points I raised in that review. Thus, I recommend acceptance after the authors address the very minor points that I noted while reading this version of the manuscript. Those points are as follows:

\textcolor{blue}{Thank you for your feedback on our manuscript. We have addressed all your comments and suggestions, and our replies to your comments can be found in blue below (original comments in black).}

+ Footnote 1 – change "saturated thickness" to "aquifer thickness" to be consistent with usage elsewhere. Same for x-axis label on Figure C.2.

\textcolor{blue}{Thank you for catching this. In the footnote, "saturated thickness" was replaced with "aquifer thickness".}

+ Line 110 – can you briefly describe how this "accounting" is done?

\textcolor{blue}{The figure is designed to showcase the effect of aquifer thickness on the percentage of irrigated acres, without the influence of spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. In deriving the relationship between aquifer thickness and the proportion of irrigated acres using our model, these factors are kept constant, allowing us to focus solely on the impact of aquifer thickness. We have opted to omit the latter part of the sentence to avoid potential confusion. While Figure 1 was constructed using a similar procedure—adjusting only the water deficit while keeping other factors unchanged — we did not explicitly state this either. We believe this omission avoids unnecessary confusion.}

+ Line 147 – "leading in" should be "leading to".

\textcolor{blue}{Thank you for catching this. In the footnote, "leading in" was replaced with "leading to".}

+ Lines 282-283 – delete "of the aquifer" - redundant.

\textcolor{blue}{Thank you for catching this. In the sentence, "of the aquifer" was removed to avoid redundancy.}

+ Line 289 – "thickness" should be "thicknesses".

\textcolor{blue}{Thank you for catching this. In the sentence, "thickness" was replaced with "thicknesses".}

+ Figure 7 – need a 200 m label for scale bar on soybean map. Are the blank areas in the aquifer in Texas and Oklahoma an indication that the aquifer is thin or that there is a lack of data? Please clarify.

+ Line 308 – "y" should be "Y".

\textcolor{blue}{Thank you for catching this. The suggested changes has been made.}

+ Lines 357-359 – delete sentence as it is repeated on lines 363-364 where it fits better.

\textcolor{blue}{Thank you for catching this. The sentenced was removed.}

+ Lines 401-402 – a recent paper in Agricultural Water Management addresses this issue in more detail 
(https://www.sciencedirect.com/science/article/pii/S0378377423002731).

\textcolor{blue}{Thank you for suggesting this article. We now cite this article at the end of the sentence.}

+ Lines 410-413 – very awkward sentence – please rewrite.

\textcolor{blue}{We have rewritten the sentence to the following:}

\textcolor{blue}{"As discussed in section 1 of our manuscript, there is extensive evidence pointing to reduced well yields in these areas of the HPA. This evidence ranges from anecdotal and theoretical insights to simulation-based studies. Collectively, they suggest that these restrictions significantly constrain farmers' ability to meet crop water demands during droughts."}

+ Line 446 – delete "across".

\textcolor{blue}{Thank you for catching this. The suggested changes has been made.}

# Reply to Reviewer 3

The second round comments have been addressed convincingly. The discussion of the role of groundwater management policies in Appendix A is particularly useful addition, as it clarifies how the results presented here relate to the findings of analyses which focused more on policy factors. Ruling out regulations as a substantial causal factor based on the limited temporal overlap with the study period is coherent. The other edits implemented in response to the other comments have further improved the strength and accuracy of the presentation.

Two relatively minor points remain to be addressed, based on one of the responses and the data and code correction you mentioned:

\textcolor{blue}{Thank you for your feedback on our manuscript. We have addressed all your comments and suggestions, and our replies to your comments can be found in blue below (original comments in black).}

+ (1) I would recommend a minor addition regarding the response to my second round comment 5.b. You have pointed out that the claim of a significant yield decline in lines 77-78 is supported by Figure A.1 (and also A.2 for soybeans). Please add a reference to both figures in this sentence to clarify for readers where they can find evidence for the statistical significance.

\textcolor{blue}{Thank you for this suggestion. We modify the sentence as follows:}

\textcolor{blue}{As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures A.1 and A.2 for the 95\% confidence intervals)}

+ (2) The beginning of the response letter describes a code and data correction unrelated to the previous reviewer comments. "[Counties] with only irrigated acres" were removed from the irrigation share regression. The immediate acknowledgement of this correction is appreciated. I trust that it was based on sensible considerations, though the rationale does not immediately become clear from the short statement, especially given that this does not seem to have been relevant previously. As you point out, the result changes due to this correction are quantitatively relatively minor (most visible in Figs. 3 and 4), and do not affect the qualitative findings or conclusions.

\textcolor{blue}{We apologize for the confusion we caused. Our sentnce on this issue was not clear in the last round of revision. We meants that we excluded counties where only the irrigated acres are "observed" because only irrigated acres were reported by USDA, but not rainfed acres. So, it is very much possible that those counties had rainfed production, but rainfed acres were simply not reported by USDA for the counties we dropped. So, we were not trying to drop the fully irrigated counties. Of course, they may be fully irrigated and actually had zero rainfed production though it is very unlikely. If we keep them, they would all have irrigated share of 1 (100\%) just because drayland acres were not repoted by USDA, which can be a significant over-estimation of their actual share of irrigated acres. Therefore, we should drop them to avoid the problem lowering the quality of the data and consequently analysis. This is more of a data limitation than anything.}

I recommend, however, making sure that the correction, its rationale, and potential implications for the result interpretation are made clear to readers. Lines 347-348 state that "irrigated area share and average crop yield estimates use data only from counties that have both irrigated and rainfed production observations". This statement covers the removal of fully irrigated counties, but could be clearer: Lines 348-350 mention the removal of "solely rainfed" counties explicitly - mentioning this but not the removal of fully irrigated counties could be confusing to readers. Please include an explicit statement about the removal of fully irrigated counties to avoid confusion. Please also make clear whether this removal pertains to both regressions (as implied in lines 347) or just the irrigated share regression (the one highlighted in the letter).

\textcolor{blue}{We modified the pertinent sentences to the following:}

\textcolor{blue}{Note that irrigated area share and average crop yield estimates use data only from counties that have both irrigated and rainfed production observations because irrigated area share cannot be calculated when missing either one of the irrigated and rainfed observations. Counties with solely irrigated or rainfed production observation data (Section 4.1) are used solely for estimating impacts of drought on per-area crop yields (Equation 1).}

Please also provide a short rationale, e.g. whether this is due to the model selected or substantive considerations. Finally, please clarify what this implies for the result interpretation. Fully irrigated county observations do not seem obviously irrelevant. The change in Fig. 3 results due to the correction is quantitatively minor but not irrelevant. It would, therefore, be helpful for readers to understand whether the exclusion of fully irrigated counties makes the results more accurate, or whether it creates a (necessary) bias.

\textcolor{blue}{Please see our replies about the rational. We simply cannot calculate irrigated area share for counties with only irrigated or ranifed observations.}

Relatedly, it seems that in Fig. 7, some of the previously displayed counties in both the irrigated (color shaded) and rainfed (grey) categories are now missing. Did the exclusion of fully irrigated counties also affect which rainfed areas are represented? If so, please also clarify this link for the readers.

\textcolor{blue}{Thank you so much for noticing this change. This has nothing to do with the exclusion of counties with only irrigation acres reported. The maps were supposed to present saturated thickness for "all" the counties used in the analysis. So, even the counties that were dropped for the irrigation share analyis should have been present. After some digging, we found that we actually made mistakes twice on map creation during the whole review process so far.}

1. \textcolor{blue}{In the 1st round, we made a correction to the lower bound of saturated thickness according to reviewer 1's comment. We stated that 9m was used as the lower bound as below.}

\textcolor{blue}{"Aquifer thickness values were only estimated for counties with irrigated yield observations. Furthermore, during the aggregation process, any aquifer cells within a county that had a aquifer thickness value lower than 9 m were removed before aggregation to reflect the fact that aquifer thickness lower than this threshold are not considered to be viable for extraction and therefore it is unlikely these areas contribute groundwater for irrigation production [Fenichel et al., 2016, Haacker et al., 2016, Deines et al., 2020]."}

\textcolor{blue}{However, we were actually using 12m as the lower bound, which we realized after seeing reviewer's comment on this in the first round:}

\textcolor{blue}{"Figure 2 - Why do the plots stop at a thickness of about 12 m when you had earlier defined the minimum thickness as 9 m and mention 10 m in the text when discussing the relationships?"}

\textcolor{blue}{Therefore, we change the code so that 9m is used instead of 12m. We, however, did not update the map according to this change in the data. This is why the map in this version has several more counties in Kansas and New Mexico that did not appear in the map of the first round revision. They used to be dropped prior to chaging the threshold from 12m to 9m.}

2. \textcolor{blue}{The significant change in the map in the second round from the 1st round is a mistake on using the map we created when we were experimenting with codes. We then did not rerun the code to create a map with the whole dataset to overwrite the wrong map with the correct map. We made sure that we streamlined all the codes, rerun all of them, and cofirm that the map indeed reflect what we used for our analysis in this round. As stated in the manuscript, all the codes and data are available on a publicly accessible Github repository. Thank you again for your keen eyes on details. We would not have noticed this ourselves.}

### 2023-10-09 Writing/NatureWater/3rdRoundRevision/replies/reply_to_reviewers_R3.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\hypertarget{reply-to-reviewer-1}{%
\section{Reply to Reviewer 1}\label{reply-to-reviewer-1}}

@@

I have read the second revision of this manuscript and the authors'
responses to my last review. I think the authors have done an excellent
job in responding to the points I raised in that review. Thus, I
recommend acceptance after the authors address the very minor points
that I noted while reading this version of the manuscript. Those points
are as follows:

\textcolor{blue}{Thank you for your feedback on our manuscript. We have addressed all your comments and suggestions, and our replies to your comments can be found in blue below (original comments in black).}

\begin{itemize}
\tightlist
\item
  Footnote 1 -- change ``saturated thickness'' to ``aquifer thickness''
  to be consistent with usage elsewhere. Same for x-axis label on Figure
  C.2.
\end{itemize}

@@

\textcolor{blue}{Thank you for catching this. In the footnote, "saturated thickness" was replaced with "aquifer thickness".}

\begin{itemize}
\tightlist
\item
  Line 110 -- can you briefly describe how this ``accounting'' is done?
\end{itemize}

@@

\textcolor{blue}{The figure is designed to showcase the effect of aquifer thickness on the percentage of irrigated acres, without the influence of spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. In deriving the relationship between aquifer thickness and the proportion of irrigated acres using our model, these factors are kept constant, allowing us to focus solely on the impact of aquifer thickness. We have opted to omit the latter part of the sentence to avoid potential confusion. While Figure 1 was constructed using a similar procedure—adjusting only the water deficit while keeping other factors unchanged — we did not explicitly state this either. We believe this omission avoids unnecessary confusion.}

\begin{itemize}
\tightlist
\item
  Line 147 -- ``leading in'' should be ``leading to''.
\end{itemize}

@@

\textcolor{blue}{Thank you for catching this. In the footnote, "leading in" was replaced with "leading to".}

\begin{itemize}
\tightlist
\item
  Lines 282-283 -- delete ``of the aquifer'' - redundant.
\end{itemize}

@@

\textcolor{blue}{Thank you for catching this. In the sentence, "of the aquifer" was removed to avoid redundancy.}

\begin{itemize}
\tightlist
\item
  Line 289 -- ``thickness'' should be ``thicknesses''.
\end{itemize}

@@

\textcolor{blue}{Thank you for catching this. In the sentence, "thickness" was replaced with "thicknesses".}

\begin{itemize}
\item
  Figure 7 -- need a 200 m label for scale bar on soybean map. Are the
  blank areas in the aquifer in Texas and Oklahoma an indication that
  the aquifer is thin or that there is a lack of data? Please clarify.
\item
  Line 308 -- ``y'' should be ``Y''.
\end{itemize}

@@

\textcolor{blue}{Thank you for catching this. The suggested changes has been made.}

\begin{itemize}
\tightlist
\item
  Lines 357-359 -- delete sentence as it is repeated on lines 363-364
  where it fits better.
\end{itemize}

@@

\textcolor{blue}{Thank you for catching this. The sentenced was removed.}

\begin{itemize}
\tightlist
\item
  Lines 401-402 -- a recent paper in Agricultural Water Management
  addresses this issue in more detail
  (\url{https://www.sciencedirect.com/science/article/pii/S0378377423002731}).
\end{itemize}

@@

\textcolor{blue}{Thank you for suggesting this article. We now cite this article at the end of the sentence.}

\begin{itemize}
\tightlist
\item
  Lines 410-413 -- very awkward sentence -- please rewrite.
\end{itemize}

@@

\textcolor{blue}{We have rewritten the sentence to the following:}

\textcolor{blue}{"As discussed in section 1 of our manuscript, there is extensive evidence pointing to reduced well yields in these areas of the HPA. This evidence ranges from anecdotal and theoretical insights to simulation-based studies. Collectively, they suggest that these restrictions significantly constrain farmers' ability to meet crop water demands during droughts."}

\begin{itemize}
\tightlist
\item
  Line 446 -- delete ``across''.
\end{itemize}

@@

\textcolor{blue}{Thank you for catching this. The suggested changes has been made.}

\hypertarget{reply-to-reviewer-3}{%
\section{Reply to Reviewer 3}\label{reply-to-reviewer-3}}

@@

The second round comments have been addressed convincingly. The
discussion of the role of groundwater management policies in Appendix A
is particularly useful addition, as it clarifies how the results
presented here relate to the findings of analyses which focused more on
policy factors. Ruling out regulations as a substantial causal factor
based on the limited temporal overlap with the study period is coherent.
The other edits implemented in response to the other comments have
further improved the strength and accuracy of the presentation.

Two relatively minor points remain to be addressed, based on one of the
responses and the data and code correction you mentioned:

\textcolor{blue}{Thank you for your feedback on our manuscript. We have addressed all your comments and suggestions, and our replies to your comments can be found in blue below (original comments in black).}

\begin{itemize}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \tightlist
  \item
    I would recommend a minor addition regarding the response to my
    second round comment 5.b. You have pointed out that the claim of a
    significant yield decline in lines 77-78 is supported by Figure A.1
    (and also A.2 for soybeans). Please add a reference to both figures
    in this sentence to clarify for readers where they can find evidence
    for the statistical significance.
  \end{enumerate}
\end{itemize}

@@

\textcolor{blue}{Thank you for this suggestion. We modify the sentence as follows:}

\textcolor{blue}{As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures A.1 and A.2 for the 95\% confidence intervals)}

\begin{itemize}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \setcounter{enumi}{1}
  \tightlist
  \item
    The beginning of the response letter describes a code and data
    correction unrelated to the previous reviewer comments.
    ``{[}Counties{]} with only irrigated acres'' were removed from the
    irrigation share regression. The immediate acknowledgement of this
    correction is appreciated. I trust that it was based on sensible
    considerations, though the rationale does not immediately become
    clear from the short statement, especially given that this does not
    seem to have been relevant previously. As you point out, the result
    changes due to this correction are quantitatively relatively minor
    (most visible in Figs. 3 and 4), and do not affect the qualitative
    findings or conclusions.
  \end{enumerate}
\end{itemize}

@@

\textcolor{blue}{We apologize for the confusion we caused. Our sentnce on this issue was not clear in the last round of revision. We meants that we excluded counties where only the irrigated acres are "observed" because only irrigated acres were reported by USDA, but not rainfed acres. So, it is very much possible that those counties had rainfed production, but rainfed acres were simply not reported by USDA for the counties we dropped. So, we were not trying to drop the fully irrigated counties. Of course, they may be fully irrigated and actually had zero rainfed production though it is very unlikely. If we keep them, they would all have irrigated share of 1 (100\%) just because drayland acres were not repoted by USDA, which can be a significant over-estimation of their actual share of irrigated acres. Therefore, we should drop them to avoid the problem lowering the quality of the data and consequently analysis. This is more of a data limitation than anything.}

I recommend, however, making sure that the correction, its rationale,
and potential implications for the result interpretation are made clear
to readers. Lines 347-348 state that ``irrigated area share and average
crop yield estimates use data only from counties that have both
irrigated and rainfed production observations''. This statement covers
the removal of fully irrigated counties, but could be clearer: Lines
348-350 mention the removal of ``solely rainfed'' counties explicitly -
mentioning this but not the removal of fully irrigated counties could be
confusing to readers. Please include an explicit statement about the
removal of fully irrigated counties to avoid confusion. Please also make
clear whether this removal pertains to both regressions (as implied in
lines 347) or just the irrigated share regression (the one highlighted
in the letter).

@@

\textcolor{blue}{We modified the pertinent sentences to the following:}

\textcolor{blue}{Note that irrigated area share and average crop yield estimates use data only from counties that have both irrigated and rainfed production observations because irrigated area share cannot be calculated when missing either one of the irrigated and rainfed observations. Counties with solely irrigated or rainfed production observation data (Section 4.1) are used solely for estimating impacts of drought on per-area crop yields (Equation 1).}

Please also provide a short rationale, e.g.~whether this is due to the
model selected or substantive considerations. Finally, please clarify
what this implies for the result interpretation. Fully irrigated county
observations do not seem obviously irrelevant. The change in Fig. 3
results due to the correction is quantitatively minor but not
irrelevant. It would, therefore, be helpful for readers to understand
whether the exclusion of fully irrigated counties makes the results more
accurate, or whether it creates a (necessary) bias.

@@

\textcolor{blue}{Please see our replies about the rational. We simply cannot calculate irrigated area share for counties with only irrigated or ranifed observations.}

Relatedly, it seems that in Fig. 7, some of the previously displayed
counties in both the irrigated (color shaded) and rainfed (grey)
categories are now missing. Did the exclusion of fully irrigated
counties also affect which rainfed areas are represented? If so, please
also clarify this link for the readers.

\textcolor{blue}{Thank you so much for noticing this change. This has nothing to do with the exclusion of counties with only irrigation acres reported. The maps were supposed to present saturated thickness for "all" the counties used in the analysis. So, even the counties that were dropped for the irrigation share analyis should have been present. After some digging, we found that we actually made mistakes twice on map creation during the whole review process so far.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textcolor{blue}{In the 1st round, we made a correction to the lower bound of saturated thickness according to reviewer 1's comment. We stated that 9m was used as the lower bound as below.}
\end{enumerate}

@@

\textcolor{blue}{However, we were actually using 12m as the lower bound, which we realized after seeing reviewer's comment on this in the first round:}

\textcolor{blue}{"Figure 2 - Why do the plots stop at a thickness of about 12 m when you had earlier defined the minimum thickness as 9 m and mention 10 m in the text when discussing the relationships?"}

\textcolor{blue}{Therefore, we change the code so that 9m is used instead of 12m. We, however, did not update the map according to this change in the data. This is why the map in this version has several more counties in Kansas and New Mexico that did not appear in the map of the first round revision. They used to be dropped prior to chaging the threshold from 12m to 9m.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  \textcolor{blue}{The significant change in the map in the second round from the 1st round is a mistake on using the map we created when we were experimenting with codes. We then did not rerun the code to create a map with the whole dataset to overwrite the wrong map with the correct map. We made sure that we streamlined all the codes, rerun all of them, and cofirm that the map indeed reflect what we used for our analysis in this round. As stated in the manuscript, all the codes and data are available on a publicly accessible Github repository. Thank you again for your keen eyes on details. We would not have noticed this ourselves.}
\end{enumerate}

\end{document}

### 2023-10-09 Writing/NatureWater/3rdRoundRevision/manuscript.tex
edits made

Two major pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}\footnote{Another pathway is through groundwater use regulation in response to low or declining saturated thickness. However, it is unlikely that this effects is captured in our analysis. Please see Appendix \ref{pathway-reg}}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B)\footnote{Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \ref{reg-conf} for estimated yield response curves with 95 percent confidence interval.}. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases. This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

@@

In the final step of our analysis, we then estimate the average crop yield as a function of aquifer thickness and seasonal water deficit. The average crop yield is estimated as the average of rainfed and irrigated yields estimated on the basis of Equation \eqref{eq:eqn1}, weighted by the estimated share of total acres that are irrigated in the county estimated using Equation \eqref{eq:eqn2}. In this combined model, aquifer thickness is a continuous variable for the irrigated area share regression component and a categorical variable for the irrigated yield regression following the format of Equations \eqref{eq:eqn2} and \eqref{eq:eqn1}, respectively. Comparing the estimated average crop yield across different levels of aquifer thickness for a given seasonal water deficit provides an indication of the effect of aquifer conditions on drought impacts, considering impacts of drought through changes in both per-area irrigated yields and the area of land irrigated. Standard errors of this analysis are estimated using bootstrap methods clustered by county. Note that irrigated area share and average crop yield estimates use  data only from counties that have both irrigated and rainfed production observations. Counties with solely rainfed production observation data (Section \ref{data-sets}) are excluded from these analyses, and are used solely for estimating impacts of drought on per-area crop yields (Equation \ref{eq:eqn1}).

@@

Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas \citep{marston2022, rimvsaite2023groundwater}, these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to only 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in Northwest Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the Sheridan 6 LEMA area covers only parts of two counties in Kansas and hence represents a very small proportion of our overall study sample. Additionally, the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Furthermore, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions \citep{drysdale2018adaptation, deines2021combining}. These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.

Finally, we also highlight that areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction. This is due to laws that typically provide farmers with absolute rights to use groundwater under their land, coupled with low levels of metering. Research has shown that these conditions have historically limited or constrained potential for implementing extensive and binding pumping restrictions on irrigators in these areas \citep{wheeler2016lessons, closas2018chronicle}, again indicating that regulations are not the binding driver of water use constraints in these areas. In contrast, as discussed in the \ref{main} section of our manuscript, there is extensive evidence highlighting the reductions in well yields that have occurred in these parts of the HPA along with anecdotal, theoretical and simulation based evidence for why these restrictions would impose meaningful restrictions on farmers ability to adequately meet crop water needs during droughts.

>>>>

Two major pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability \citep{foster2015analysis}\footnote{Another pathway is through groundwater use regulation in response to low or declining aquifer thickness. However, it is unlikely that this effects is captured in our analysis. Please see Appendix \ref{pathway-reg}}. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs \citep{mieno2017price, bhattarai2021impact}. Second, depletion also reduces the thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation \citep{konikow2005groundwater, foster2014modeling, hrozencik2017heterogeneous}. If the reduction in well yields is substantial enough - for example as would occur when a large proportion of aquifer thickness is depleted \citep{hecox2002calculation, korus2020depletion} - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production \citep{foster2014modeling, rad2020effects}. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change \citep{ukkola2020robust, chiang2021evidence, cook2020twenty}.

@@

Figure \ref{fig:yield-response} shows the estimated crop yield response to water deficit for rainfed production and groundwater-irrigated production at different levels of aquifer thickness categories for corn (panel A) and soybean (panel B)\footnote{Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \ref{reg-conf} for estimated yield response curves with 95 percent confidence interval.}. The histogram of annual water deficit is presented at the top for each crop. In wetter years with low crop irrigation requirements (i.e., negative values of water deficit), rainfed and irrigated yields are very similar for both corn and soybean. As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures \ref{fig:irrigated-yield-ind-corn} and \ref{fig:irrigated-yield-ind-soy} for the 95\% confidence intervals). This result reflects the positive benefits irrigation provides both as a buffer against precipitation variability, and as a means of raising overall yields through increased evapotranspiration and associated biomass accumulation, reductions of heat-related stressors, and other factors \citep{zhu2022untangling, li2020quantifying}.

@@

In the final step of our analysis, we then estimate the average crop yield as a function of aquifer thickness and seasonal water deficit. The average crop yield is estimated as the average of rainfed and irrigated yields estimated on the basis of Equation \eqref{eq:eqn1}, weighted by the estimated share of total acres that are irrigated in the county estimated using Equation \eqref{eq:eqn2}. In this combined model, aquifer thickness is a continuous variable for the irrigated area share regression component and a categorical variable for the irrigated yield regression following the format of Equations \eqref{eq:eqn2} and \eqref{eq:eqn1}, respectively. Comparing the estimated average crop yield across different levels of aquifer thickness for a given seasonal water deficit provides an indication of the effect of aquifer conditions on drought impacts, considering impacts of drought through changes in both per-area irrigated yields and the area of land irrigated. Standard errors of this analysis are estimated using bootstrap methods clustered by county. Note that irrigated area share and average crop yield estimates use data only from counties that have both irrigated and rainfed production observations because irrigated area share cannot be calculated when missing either one of the irrigated and rainfed observations. Counties with solely irrigated or rainfed production observation data (Section \ref{data-sets}) are used solely for estimating impacts of drought on per-area crop yields (Equation \ref{eq:eqn1}).

@@

Where more restrictive groundwater withdrawal limits have been introduced, for example in some Groundwater Management Districts in Kansas \citep{marston2022, rimvsaite2023groundwater}, these have universally been implemented in the last 5-10 years and therefore have no or only minimal overlap with our study period which extends up to only 2016. The earliest of these groundwater managements in Kansas interventions was the Sheridan 6 Local Enhanced Management Area (LEMA) in Northwest Kansas, which introduced restrictions to farmer groundwater pumping of 11 inches per year on average per 5-year period starting in 2013. While this overlaps with the final 4 years of our study period, the Sheridan 6 LEMA area covers only parts of two counties in Kansas and hence represents a very small proportion of our overall study sample. Additionally, the years 2013-2016 were all characterized by average or above average growing season rainfall (i.e. not drought years). Furthermore, evidence shows that the effect of restrictions introduced by the LEMA has primarily been a reduction in per-area water use, with irrigated crop yields largely unaffected and only small reductions in irrigated areas compared to pre-policy conditions \citep{drysdale2018adaptation, deines2021combining, whittemore2023108408}. These policies therefore would not be expected to confound our results on the impacts of aquifer thickness.

Finally, we also highlight that areas of the southern HPA, in particular in northern Texas, where the lowest aquifer thickness values are observed have few if any regulatory restrictions on groundwater extraction. This is due to laws that typically provide farmers with absolute rights to use groundwater under their land, coupled with low levels of metering. Research has shown that these conditions have historically limited or constrained potential for implementing extensive and binding pumping restrictions on irrigators in these areas \citep{wheeler2016lessons, closas2018chronicle}, again indicating that regulations are not the binding driver of water use constraints in these areas. As discussed in section \ref{main} of our manuscript, there is extensive evidence pointing to reduced well yields in these areas of the HPA. This evidence ranges from anecdotal and theoretical insights to simulation-based studies. Collectively, they suggest that these restrictions significantly constrain farmers' ability to meet crop water demands during droughts.
